---
title: "Practical 2"
params: 
    SHOW_SOLS: FALSE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
source('assets/setup.R')
library(tidyverse)
qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
library(xaringanExtra)
use_panelset()

library(DasGuptR)
```

# Example: Reconvictions by age group

TODO 

# Example 2: Reconvictions by age and sex 

TODO

# choose your own


::: {.callout-note collapse="true"}
#### Scotland's Homelessness Deaths

Scotland tracks the numbers of homeless deaths each year.  
It also tracks the number of homelessness applications each year.  
Both of these measures are provided broken down into four coarse age groups.  

One way in which we could choose to approach this question is to decompose the number of homeless deaths per 100,000 people in the population:  

$$ \small
\begin{align}
\text{Homelessness Death Rate} &= \sum\limits_{\text{agegroup }i} \frac{\text{Nr Deaths}_i}{\text{Nr People}_i} \cdot 100,000\\
&= \sum\limits_{\text{agegroup }i}\frac{\text{Nr Applications}_i}{\text{Nr People}_i} \cdot \frac{\text{Nr Deaths}_i}{\text{Nr Applications}_i} \cdot 100,000 \\
\end{align}
$$

There is an important note here that the number of applications is a measure that doesn't represent the entire homeless population, but rather those that are currently engaging with the system in a given year. As such it captures the mortality risk of the population that is engaging with applications, and *not* the mortality risk of the entire homeless population.  

An alternative approach we might choose here is to decompose the *total* number of homeless deaths. The benefit here is that we can include the raw population size as a contributing factor:  

$$ \small
\text{Total Homelessness Deaths} = \sum\limits_{\text{agegroup }i} \text{Nr People} \cdot \frac{\text{Nr People}_i}{\text{Nr People}} \cdot \frac{\text{Nr Applications}_i}{\text{Nr People}_i} \cdot \frac{\text{Nr Deaths}_i}{\text{Nr Applications}_i} \\
$$

If our conversation is focussed on the raw totals (homeless deaths are up from 164 in 2017 to 231 in 2024), this may be preferable as we can provide the % due to population growth, the % due to an aging demographic, the % due to changes in applications, and the % due to the change in mortality risk per application.  

Data are available at [https://josiahpjking.github.io/sgsss_std_decomp/data/hdeaths.csv](https://josiahpjking.github.io/sgsss_std_decomp/data/hdeaths.csv){target="_blank"}  

:::
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

decomposing deaths per 100,000 people:  
```{r}
hd <- read_csv("data/hdeaths.csv")
hd$scotpop = sample(hd$scotpop)
hd <- 
  hd |> mutate(
    hdr = deaths/applications,
    hr = applications/scotpop,
  ) 
  
#crude rates
hd |> 
  group_by(year) |>
    summarise(
      cr = sum(deaths)/sum(scotpop)*1e5,
      # as multiplication of factors: 
      cr2 = sum( hdr * hr * scotpop/sum(scotpop) )*1e5
    )
  

dgres <- 
  dgnpop(hd, 
         pop="year",
         factors=c("hr","hdr"),
         id_vars="age_group",
         crossclassified = "scotpop",
         ratefunction = "hr*hdr*1e5")

dg_plot(dgres, legend.position="bottomright")
dgres |> dg_table(2017,2024)

```

because there is just one sub-population grouping here, we could alternatively do: 
```{r}
#| eval: false
hd <- hd |>
  group_by(year) |>
  mutate(
    popshare = scotpop/sum(scotpop)
  )

dgres.a <- 
  dgnpop(hd, 
         pop="year",
         factors=c("hr","hdr","popshare"),
         ratefunction = "sum(popshare*hr*hdr*1e5)")

dg_plot(dgres.a, legend.position="bottomright")
```


decomposing total deaths: 
```{r}
hd <- hd |> group_by(year) |>
  mutate(
    tpop = sum(scotpop)
  )


dgres2 <- 
  dgnpop(hd, 
         pop="year",
         factors=c("hr","hdr","tpop"),
         id_vars="age_group",
         crossclassified = "scotpop")

dg_plot(dgres2, legend.position="bottomright")
dgres |> dg_table(2017,2024)
```



`r solend()`






::: {.callout-note collapse="true"}
#### Scotland's Recycling

In the first workshop, one example saw the decomposition of the "kg recycled waste per person":  
$$
\text{Recycled Tonnes per Person} = \frac{\text{Total Waste}}{\text{Nr People}} \cdot \frac{\text{Recycled Waste}}{\text{Total Waste}}
$$

But we could also consider decomposing the total tonnage that is recycled, and include the population size as another factor:  
$$
\text{Recycled Tonnes} = \text{Nr People} \cdot \frac{\text{Total Waste}}{\text{Nr People}} \cdot \frac{\text{Recycled Waste}}{\text{Total Waste}}
$$

SEPA provides data on this every year from 2011 onwards, providing us with a good example for decomposition across a time series.  

The data are available at [https://josiahpjking.github.io/sgsss_std_decomp/data/recycling_timeseries.csv](https://josiahpjking.github.io/sgsss_std_decomp/data/recycling_timeseries.csv){target="_blank"}  

:::

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
df <- read_csv("data/recycling_timeseries.csv")

df <- df |>
  mutate(
    wpp = totalwaste/pop,
    recp = recycled/totalwaste,
    lfp = landfilled/totalwaste,
    incp = diverted/totalwaste
  )

# recycling tonnes per person:
dgnpop(df,
       pop="year",
       factors=c("wpp","recp")) |> 
  dg_plot(legend.position="bottom")

# recycling tonnage
dgnpop(df,
       pop="year",
       factors=c("pop","wpp","recp")) |> 
  dg_plot(legend.position="bottom")


```

`r solend()`



::: {.callout-note collapse="true"}
#### age-specific death rates in POP1 and POP2

two countries age-specific death rates (deaths / total)  
Rate = sum( deaths_g / total_g )  
age-specific deaths will probs be found separately from age-specific population estimates

:::



::: {.callout-note collapse="true"}
#### Scotland's Business 5 year Survival Rate

Rate = survivey1/births * survivey3/survivey1 * survivey5/survivey3

do it across sectors i

:::




::: {.callout-note collapse="true"}
#### Scotland's Heroin Deaths 2020 vs 2024





:::






