---
title: "workshop-1"
---

# Find your own!  

death rates across council areas

https://www.nrscotland.gov.uk/publications/deaths-time-series-data-2024/


1. pick an age group.
2. pick two LAAs
3. how many deaths in that age group in that year for each LAA?
4. how many people were in that age group that year for each LAA?
5. what is the crude death rate for each LAA? (ndeaths/total pop)
6. what would the death rate for each LAA be if you held the proportion of that age group at the average? 
7. how much of the difference in crude rates can be attributed to this compositional difference between LAAs? 


EXAMPLE:  
```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)

scotlaa <- tribble(
  ~laa, ~n_people, ~n_over60, ~deaths_over60,
  "Glasgow City", 650300, 128902, 5287,
  #"Highland", 237290, 77271, 2564,
  #"East Dunbartonshire", 109970, 35532, 1114,
  "Edinburgh", 530680, 113979, 3877
)

rbind(as.data.frame(scotlaa), "...","...") |>
  gt::gt()
```

crude rates
and showing that r = a*b
```{r}
scotlaa |>
  mutate(
    cruderate = deaths_over60/n_people*1e3,
    prop_over60 = n_over60/n_people,
    over60_deathrate = deaths_over60 / n_over60,
    cruderate2 = prop_over60 * over60_deathrate*1e3
  )
```
    
std rates
```{r}
scotlaa |>
  mutate(
    avgprop = mean(n_over60/n_people),
    avgover60deathrate = mean(deaths_over60 / n_over60),
    stdrate1 = avgprop* (deaths_over60 / n_over60) * 1e3,
    stdrate2 = (n_over60/n_people) * avgover60deathrate
  )
```
    
decomp
```{r}
scotlaa |>
  mutate(
    avgprop = mean(n_over60/n_people),
    avgover60deathrate = mean(deaths_over60 / n_over60),
    stdrate1 = avgprop* (deaths_over60 / n_over60) * 1e3,
    stdrate2 = (n_over60/n_people) * avgover60deathrate * 1e3,
    cr = deaths_over60/n_people*1e3
  ) |> select(stdrate1,stdrate2,cr) |>
  apply(2,diff)
```

of the crude rate difference (Edinburgh is 0.82 deaths per 1000 lower than glasgow), 175% this difference would happen if they only differed in the 'over60s death rate'.

-75% of the difference would happen if they only differed in the age demographic. i.e., Edinburgh would have a higher death rate than glasgow! 

- -1.4456220 / -0.8243728 * 100 = 175  
- 0.6212492 / -0.8243728 *100 = -75  

```{r}
library(DasGuptR)
scotlaa |>
  mutate(
    prop_over60 = n_over60/n_people,
    over60_deathrate = deaths_over60 / n_over60,
  ) |>
  dgnpop(
    pop="laa",
    factors=c("prop_over60","over60_deathrate"),
    ratefunction="prop_over60*over60_deathrate*1e3"
  ) |> dg_table()
```



