---
title: "Practical 1"
params: 
    SHOW_SOLS: FALSE
    TOGGLE: TRUE
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
source('assets/setup.R')
library(tidyverse)
qcounter <- function(){
  if(!exists("qcounter_i")){
    qcounter_i <<- 1
  }else{
    qcounter_i <<- qcounter_i + 1
  }
  qcounter_i
}
library(xaringanExtra)
use_panelset()
```


# Example: Birth Rates in Stirling vs Glasgow

```{r}
#| include: false
pop <- read_csv("data/pop_midyear_age_sex_council_areas.csv") |> 
  janitor::clean_names() |>
  filter(area_name!="Scotland") |>
  pivot_longer(x0:x90_and_over) |>
  mutate(age = parse_number(name)) |>
  group_by(year, area_name, area_code, sex) |>
  summarise(
    x15_44 = sum(value[age>=15 & age<=44]),
    others = sum(value[age<15|age>44]) 
  ) |>
  filter(sex != "Persons") |>
  filter(year == 2024) |>
  group_by(area_name) |>
  summarise(
    nw1544 = sum(x15_44[sex=="Females"]),
    pop = sum(x15_44)+sum(others)
  )

birth <- read_csv("data/births_mage_council_areas.csv") |>
  janitor::clean_names() |>
  filter(year == 2024) |> 
  transmute(
    area_name, nbirths = all_ages
  )

p01 <- left_join(pop,birth) |>
  transmute(
    LAA = area_name, n_people = pop, n_women15to44 = nw1544, n_births = nbirths
  ) |> filter(grepl("Glasg|Stirl",LAA))
```

`r qbegin(qcounter())`
Here is the data for two Local Authority Areas (LAA) in Scotland: Stirling and Glasgow City.  

The crude birth rate is calculated as the number of births per 1000 people.  
$$
\text{Birth Rate} = \frac{\text{Number of Births}}{\text{Number of People}} \cdot 1000
$$


Calculate the birth rate for each LAA

```{r}
#| echo: false
p01 |>
  select(-n_women15to44) |>
  mutate(
    birth_rate = "?"
  ) |> gt::gt()
```



`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
#| echo: false
p01 |> select(-n_women15to44) |>
  mutate(
    birth_rate = round(n_births/n_people*1000,2)
  ) |> gt::gt()
```

`r solend()`

`r qbegin(qcounter())`
We can decompose the birth rate into two components: **A:** the number of births per "woman of child bearing age "(Scotland uses ages 15-44)", and **B:** the proportion of the population that are women aged 15-44.  


$$
\text{Birth Rate} = A \cdot B \cdot 1000
$$
<br>
$$
\text{Birth Rate} = \frac{\text{Nr of Women 15-44}}{\text{Nr of People}} \cdot \frac{\text{Nr of Births}}{\text{Nr of Women 15-44}} \cdot 1000
$$

Calculate these below, and then multiply them to demonstrate to yourselves that this returns the same rates (give or take some rounding!) as the previous question.  

```{r}
#| echo: false
p01 |>
  mutate(
    `w15-44 / people` = c("A1","A2"),
    `births / w15-44` = c("B1","B2"),
    birth_rate = c("A1 * B1 * 1000", "A2 * B2 * 1000")
  ) |> gt::gt()
```


`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
#| echo: false
p01 |>
  mutate(
    `w15-44 / people` = round(n_women15to44/n_people,5),
    `births / w15-44` = round(n_births/n_women15to44,5),
    birth_rate = round(`w15-44 / people`*`births / w15-44`*1000,2)
  ) |> gt::gt()
```

`r solend()`



`r qbegin(qcounter())`
What would Stirling's birth rate be if:  

1. Women aged 15-44 made up the same proportion of its population as they do for Glasgow?

::: {.callout-tip collapse="true"}
#### Hints

A1 * B2!  


```{r}
#| echo: false
p01 |>
  transmute(
    LAA,
    `w15-44 / people` = c("A1","A2"),
    `births / w15-44` = c("B1","B2"),
  ) |> gt::gt()
```



:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
p01 <- p01 |>
  mutate(
    bpw = round(n_births/n_women15to44,5),
    propw = round(n_women15to44/n_people,5),
    birth_rate = bpw*propw*1000
  )

p01
```

```{r}
p01$propw[1] * p01$bpw[2] * 1000
```


`r solend()`


`r qbegin(qcounter())`
What would each LAA's birth rates be if:  

- The proportion of the population comprised of women aged 15-44 was the same in both populations (i.e., at the average of the two)
- The rates of births in women aged 15-44 were the same in both populations (i.e., at the average of the two)

::: {.callout-tip collapse="true"}
#### Hints

B1 * ((A1+A2)/2)  
B2 * ((A1+A2)/2)  
and  
A1 * ((B1+B2)/2)  
A2 * ((B1+B2)/2)  

```{r}
#| echo: false
p01 |>
  transmute(
    LAA,
    `w15-44 / people` = c("A1","A2"),
    `births / w15-44` = c("B1","B2"),
  ) |> gt::gt()
```

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`


```{r}
p01

# if proportion of w15-44 were equal:
p01$bpw * mean(p01$propw) * 1000
# if births per w15-44 were equal:
mean(p01$bpw) * p01$propw * 1000
```


`r solend()`

`r qbegin(qcounter())`
The differences between these two pairs of standardised rates (when using the average of the two populations as the reference) provides the contribution of each factor on the total crude rate difference. 

We know from the very first question that Stirling has a crude birth rate that is ~2.46 births per 1000 people fewer than Glasgow.  

How much of this is due to Stirling having relatively fewer 'women of child bearing age'?  

::: {.callout-tip collapse="true"}
#### Hints

(A2 - A1) * ((B1+B2)/2)  / (R2 - R1)

```{r}
#| echo: false
p01 |>
  transmute(
    LAA,
    `crude rate` = c("R1","R2"),
    `w15-44 / people` = c("A1","A2"),
    `births / w15-44` = c("B1","B2"),
  ) |> gt::gt()
```

:::

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
# these are Glasgow and Stirling's propw-standardised rates, and births-per-women standardised rates: 
p01 <- p01 |> mutate(
  pw.std = bpw * mean(propw) * 1000,
  br.std = mean(bpw) * propw * 1000
)
p01

# 89% of the difference in rates is due to differences in proportions of women 15-44!  
( p01$br.std[2] - p01$br.std[1] ) / 
  (p01$birth_rate[2] - p01$birth_rate[1] ) * 100
```

`r solend()`


# Get The Software

`r qbegin(qcounter())`
Get either R or Stata open, and install the various packages. 

In R: **DasGuptR**    
We would recommend installing the development version
```{r}
#| eval: false
library(remotes)
install_github("josiahpjking/DasGuptR@develop")
```

In Stata: **rdecompose**

```
clear
input str12 LAA long n_people long n_women15to44 int n_births
"Glasgow City" 650300 161105 6150
"Stirling" 94210 17844 659
end
```


`r qend()`

`r qbegin(qcounter())`
Here's the data:  

::::panelset
:::panel
#### R
```{r}
p01 <- data.frame(
  LAA = c("Glasgow City","Stirling"),
  n_people = c(650300,94210),
  n_women15to44 = c(161105,17844),
  n_births = c(6150,659)
)
```
:::
:::panel
#### Stata

```
clear
input str12 LAA long n_people long n_women15to44 int n_births
"Glasgow City" 650300 161105 6150
"Stirling" 94210 17844 659
end
```
:::
::::

<br>
Create the new factors, and use either `dgnpop()` (in R) or `rdecompose` (Stata) to conduct the decomposition.  

Check that it matches your calculations above.  

You might find it easiest to multiply the "births per woman aged 15-44" by 1000 when you create it.  

In R, this will look something like this:  
```{r}
#| eval: false
library(tidyverse)
library(DasGuptR)
p01 <- p01 |>
  mutate(
    propw =  ????
    bpw =  ???
  )

dres <- dgnpop(x = ???, 
       pop = "???",
       factors = c("???","???"))

dg_table(dres)
```

`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
library(tidyverse)
library(DasGuptR)
p01 <- p01 |>
  mutate(
    propw =  n_women15to44/n_people,
    bpw =  n_births/n_women15to44*1000
  )

dres <- dgnpop(x = p01, 
       pop = "LAA",
       factors = c("bpw","propw"))

dg_table(dres)
```
`r solend()`

`r qbegin(qcounter())`
We can split the birth rate further, into three different factors:  

$$ \small
\text{Birth Rate} = \frac{\text{Nr of Women}}{\text{Nr of People}} \cdot \frac{\text{Nr of Women 15-44}}{\text{Nr of Women}} \cdot \frac{\text{Nr of Births}}{\text{Nr of Women 15-44}} \cdot 1000
$$

Perform this decomposition now.  


::::panelset
:::panel
### R
```{r}
p01a <- data.frame(
  LAA = c("Glasgow City","Stirling"),
  n_people = c(650300,94210),
  n_women = c(332054, 48731),
  n_women15to44 = c(161105,17844),
  n_births = c(6150,659)
)
```
:::
:::panel
### Stata
```default
clear
input str12 LAA long n_people long n_women long n_women15to44 int n_births
"Glasgow City" 650300 332054 161105 6150
"Stirling"     94210  48731  17844  659
end
```
:::
::::
`r qend()`
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
p01a <- p01a |> mutate(
  propw = n_women/n_people,
  propw1544 = n_women15to44/n_women,
  bpw = n_births/n_women15to44*1000
)

dgnpop(p01a,
       pop="LAA",
       factors=c("propw","propw1544","bpw")) |>
  dg_table()
```


`r solend()`


# Choose your own

::: {.callout-note collapse="true"}
#### Birth Rates Across England and Scotland

Just the same as above, but comparing England and Scotland. 

$$ \small
\text{Birth Rate} = \frac{\text{Nr of Women}}{\text{Nr of People}} \cdot \frac{\text{Nr of Women 15-44}}{\text{Nr of Women}} \cdot \frac{\text{Nr of Births}}{\text{Nr of Women 15-44}} \cdot 1000
$$

__Data:__  

::::panelset
:::panel
#### R
```{r}
scot_v_eng <- 
  tribble(
    ~country, ~n_people, ~n_women, ~n_women1544, ~n_births,
    "Scotland", 5546900, 2851740, 1064441, 45763,
    "England", 58620101, 29895762, 11583779, 567708
  )
```
:::
:::panel
#### Stata

```default
clear
input str8 country long n_people long n_women long n_women1544 long n_births
"Scotland"  5546900  2851740  1064441  45763
"England"   58620101 29895762 11583779 567708
end
```
:::
::::
:::
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`

```{r}
scot_v_eng <- 
  tribble(
    ~country, ~n_people, ~n_women, ~n_women1544, ~n_births,
    "Scotland", 5546900, 2851740, 1064441, 45763,
    "England", 58620101, 29895762, 11583779, 567708
  )

# crude rates
scot_v_eng |>
  transmute(
    country, 
    cr = n_births/n_people*1e3
  )

# create factors:
scot_v_eng <- 
  scot_v_eng |>
  mutate(
    propw = n_women/n_people,
    propw1544 = n_women1544/n_women,
    bpw = n_births/n_women1544
  )

# decomposition
dgnpop(scot_v_eng, 
       pop = "country",
       factors = c("propw","propw1544","bpw")) |>
  dg_table()
```
`r solend()`

::: {.callout-note collapse="true"}
#### The Gender Earnings-Gap

In 2025 in the UK, the estimated weekly earnings per working aged Males was £179. The weekly earnings per working aged Females was only £7. 

These numbers can be calculated as simply: 

$$
\text{weekly £ per working aged person} = \frac{\text{total weekly earnings}}{\text{number of working aged people}}
$$

This difference could be attributable to:  

- a greater number of males in the "labour force" (people who are employed or actively seeking employent. this excludes students etc).
- higher rates of employment for males in the labour force compared to females in the labour force (employment bias)
- males working more hours per week compared to females (i.e., fewer part time roles)
- a higher average hourly wage for males compared to females (a pay bias - this is the typically "gender pay gap", unadjusted for specific roles)  

So we can decompose the total weekly earnings per person as:  

$$
\text{weekly £ per working aged person} = A \cdot B \cdot C \cdot D
$$

Where 

- A = proportion of working aged population in the labour force
- B = the proportion of the labour force that are employed
- C = the average hours worked per employee
- D = the average earnings per hour

You can find this data for a whole load of different countries at [https://ilostat.ilo.org/data/?cat_mode=subject](https://ilostat.ilo.org/data/?cat_mode=subject){target="_blank"}, if you would like to do this for somewhere else. 

::::panelset
:::panel
#### R
```{r}
work25 <- data.frame(
  pop = c("males","females"),
  working_age = c(23718900, 24613110),
  labour_force = c(18442860, 5767405),
  employees = c(4823169, 280557),
  avg_hours_pp = c(37.9, 31.7),
  total_weekly_earnings = c(4240916038, 178762504)
)
```
:::
:::panel
#### Stata
```default
clear
input str7 pop long working_age long labour_force long employees float avg_hours_pp double total_weekly_earnings
"males"   23718900 18442860 4823169 37.9 4240916038
"females" 24613110 5767405  280557  31.7 178762504
end
```
:::
::::
:::
`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
# crude rates
work25 |> transmute(
  pop,
  weeklyearnings_perperson = total_weekly_earnings/working_age
)

# create factors
work25 <- 
  work25 |> 
  mutate(
    participation = labour_force / working_age,
    employment = employees / labour_force,
    hours = avg_hours_pp,
    earnings = total_weekly_earnings/hours/employees
  )

# decompose
dgnpop(work25, 
       pop="pop",
       factors=c("participation","employment","hours","earnings")) |>
  dg_table()
```

I guess it's good that a pay-bias doesn't account for much of this difference (only ~5%).  
It's worrying that the employment-bias account for almost half!  

This is a good example to note that these methods are not going to give us evidence of a bias being present (or absent).  

All we are saying is that the different employment rates for men vs women contributes 49% of the total earnings disparity.
That is not evidence of a bias against employing women - such a bias happens at the unit level, not at the level of the population. 
In this way, Das Gupta's decomposition is essentially an act of book-keeping. The counterfactuals we get from this method (A, below) are not the same as those that we need to think about to evidence an gender employment bias (B, below). 

- **A:** "if Fs and Ms had same proportions of people in the labour force, worked the same hours, and had the same hourly wage, but differed in their employment rates in the way that they do, what would their weekly earnings be? "
- **B:** "if a given individual F was perceived to be M, does their probability of employment change (all else being equal)?"

`r solend()`


::: {.callout-note collapse="true"}
#### Scotland's Recycling: 2011 and 2022


The Scottish Environment Protection Agency (SEPA) provides [data](https://www.sepa.org.uk/environment/waste/waste-data/waste-data-reporting/household-waste-data/){target="_blank"} on the weights of waste that is generated, recycled, landfilled etc.  

One indicator tracked by many countries and organisations is the amount of recycled waste per citizen, i.e.:  
$$
\text{Recycled Tonnes Per Person} = \frac{\text{Recycled Waste}}{\text{Nr People}}
$$

In Scotland, this indicator has actually **decreased** from 195kg per person in 2011 to 186kg per person in 2022. On face value this could be interpreted as an indication that we are, as a nation, recycling less!  

But we can decompose the indicator into the amount of waste *generated* per person, and the proportion of waste that is recycled.  

This decomposition will allow us to see how much of the decrease in the population indicator can be attributable to changes in the amounts of waste generated, or in the proportion of waste that is recycled.

$$
\text{Recycled Tonnes Per Person} = \frac{\text{Total Waste}}{\text{Nr People}} \cdot \frac{\text{Recycled Waste}}{\text{Total Waste}}
$$

Data are available at [https://josiahpjking.github.io/sgsss_std_decomp/data/recycling_2011_2022.csv](https://josiahpjking.github.io/sgsss_std_decomp/data/recycling_2011_2022.csv){target="_blank"}  

:::

`r solbegin(show=params$SHOW_SOLS, toggle=params$TOGGLE)`
```{r}
df <- read_csv("https://josiahpjking.github.io/sgsss_std_decomp/data/recycling_2011_2022.csv")

# crude rates and factors
df <- df |>
  mutate(
    # crude rate
    rpp = recycled/pop*1e3,
    # factors
    wpp = totalwaste/pop*1e3,
    recp = recycled/totalwaste,
  )

df

dgnpop(df,
       pop="year",
       factors=c("wpp","recp")) |> 
  dg_table()
```

This example is a good one for figuring out directions etc.  

We get from this: 

1. the crude indicator of "kg recycled per person" goes down
2. if the waste generated per person stayed the same, but we only change the proportion of waste that is recycled, then "kg recycled per person" would have gone up. 
    - this is good! it means we're recycling a greater proportion of waste! 
3. if the proportion of waste that is recycled stays the same, but we only change the amount of waste generated per person, then "kg recycled per person" would have gone down about ~3 times as much as it did. 
    - this is also good! it suggests we're creating a lot less waste!  
    
together, we have this antagonistic relationship between these two factors and the eventual outcome. A population can make the indicator go up either by generating loads more waste (a bad thing) or by recycling a greater proportion of the waste (a good thing)! 

`r solend()`



