---
title: "Untitled"
---

## decomposing a disparity: Das Gupta

$$Y^d \perp D \mid g, x, \forall d, g, x$$



## baseline


```{r}
df <- tibble(
  C = rnorm(200), 
  G = rep(0:1,e=100),
  D = rep(0:1,100),
  ptd = rnorm(200,2,1), # individual D treatment effects
  Y = rnorm(200,G+ptd*D,1),
  Yb = rbinom(200,1,plogis(scale(G+ptd*D)))
) 
df1 = df |> mutate(
  X = factor(D),
  Group = factor(G,labels=letters[1:2])
)

```

group G disparity in Y:
```{r}
lm(Y~Group,df1) |> summary() |> pluck(coefficients)
lm(Y~Group,df1) |> plot_predictions(by=c("Group"))
```

not because of more D:
```{r}
with(df1, table(Group, X))
```

not because of D having different effects:  
```{r}
m = lm(Y~X,df1[df1$Group=="a", ])
m1 = lm(Y~X,df1[df1$Group=="b", ])
library(patchwork)
(plot_predictions(m, by=c("X")) + labs(title="group A") ) +
  (plot_predictions(m1, by=c("X")) + labs(title="group B") )
```


## prevalence

no difference in G that isn't because of D.  
it's just that there's more D1 in G1 than in G0
```{r}
df <- tibble(
  C = rnorm(200), 
  G = rep(0:1,e=100),
  D = rep(c(0,1,0,1),times=c(50,50,20,80)), # more D in G1
  umC = rnorm(200,2,1),
  ptd = rnorm(200,2+umC,1), # individual D treatment effects
  Y = rnorm(200,umC+ptd*D,1),
  Yb = rbinom(200,1,plogis(scale(ptd*D)))
) 

df1 = df |> mutate(
  X = factor(D),
  Group = factor(G,labels=letters[1:2])
)
```

group G disparity in Y:
```{r}
lm(Y~Group,df1) |> plot_predictions(by=c("Group"))
```

equal X in both groups
```{r}
with(df1, table(Group,X))
```

clue here is in the SEs. X has an effect on Y, and there's more X in groupB, resulting in group disparity in Y.  
```{r}
m = lm(Y~X,df1[df1$Group=="a", ])
m1 = lm(Y~X,df1[df1$Group=="b", ])
(plot_predictions(m, by=c("X")) + labs(title="group A") ) +
  (plot_predictions(m1, by=c("X")) + labs(title="group B") )
```

group difference disappears after removing effect of X
```{r}
lm(Y~Group,df1) |> summary()
lm(Y~X+Group,df1) |> summary()
```


## effects

```{r}
set.seed(44)
df <- tibble(
  G = rep(0:1,e=100),
  D = rep(c(0,1,0,1),times=c(50,50,50,50)), # equal D in groups
  ptd = rnorm(200,2,1), # individual D treatment effects
  Y = rnorm(200,ptd*D*(G==1),1), # effect of D is only in group 1
  Yb = rbinom(200,1,plogis(scale(ptd*D)))
) 

df1 = df |> mutate(
  X = factor(D),
  Group = factor(G,labels=letters[1:2])
)
```

group G disparity in Y:
```{r}
lm(Y~Group,df1) |> plot_predictions(by=c("Group"))
```

equal X in both groups. 
```{r}
with(df1, table(Group,X))
```

X has an effect in group B but not in group A
```{r}
m = lm(Y~X,df1[df1$Group=="a", ])
m1 = lm(Y~X,df1[df1$Group=="b", ])
(plot_predictions(m, by=c("X")) + labs(title="group A") ) +
  (plot_predictions(m1, by=c("X")) + labs(title="group B") )
```

we're estimating Y~X, so this is assumed no unmeasured confounds. 
```{r}
set.seed(44)
df <- tibble(
  G = rep(0:1,e=100),
  D = rep(c(0,1,0,1),times=c(50,50,50,50)), # equal D in groups
  umC = rnorm(200,4,1),
  ptd = rnorm(200,2+umC,1), # individual D treatment effects
  Y = rnorm(200,umC+ptd*D*(G==1),1), # effect of D is only in group 1
  Yb = rbinom(200,1,plogis(scale(ptd*D)))
) 
```


## selection

```{r}
df <- tibble(
  G = rep(0:1,e=100),
  C = rnorm(200,1,1),
  ptd = rnorm(200,2*C,1), # individual D treatment effects
) |> group_by(G) |>
  mutate(ptd2 = cut(ptd,2,labels=F)) |>
  ungroup() |>
  mutate(
    D = case_when(
      G==0 & ptd2 == 1 ~ 1,
      G==0 & ptd2 == 2 ~ 0,
      G==1 & ptd2 == 1 ~ 0,
      G==1 & ptd2 == 2 ~ 1
    ),
    Y = rnorm(200,ptd*D,1), # effect of D is only in group 1
    Yb = rbinom(200,1,plogis(scale(ptd*D)))
  ) 

df1 = df |> mutate(
  X = factor(D),
  Group = factor(G,labels=letters[1:2])
)
```

group G disparity in Y:
```{r}
lm(Y~Group,df1) |> plot_predictions(by=c("Group"))
```

equal X in both groups. 
```{r}
with(df1, table(Group,X))
```

X has an effect in group B but not in group A
```{r}
m = lm(Y~X,df1[df1$Group=="a", ])
m1 = lm(Y~X,df1[df1$Group=="b", ])
(plot_predictions(m, by=c("X")) + labs(title="group A") ) +
  (plot_predictions(m1, by=c("X")) + labs(title="group B") )
```

```{r}
library(cdgd)
results0 <- cdgd0_pa(Y="Y",D="D",G="G",
                     X="C",
                     data=df,
                     alpha=0.05)
results0$results
```


recent work: elwert yu

references & tutorials






    - $\mathbb{E}[Y|X=x] \,\,\,\text{vs}\,\,\, \mathbb{E}[Y|X=x+1]$ - for each datazone what would we expect the number of births to do if we increased the number of tescos by 1.  
    - $\mathbb{E}[Y|X=0] \,\,\,\text{vs}\,\,\, \mathbb{E}[Y|X=1]$ - for each person, what would we expect probability of birth to be if we they had a tesco (vs if they didn't)  
        
- moves us into the realms of potential outcomes (and all of the challenges that come with that!)


## regression decomposition methods: 

- Oaxaca-Blinder

- Mediation

- Others 

